\input{tlca07-talk.pre}

\pstrSetArrowColor{black}

\title{\texorpdfstring{The Safe $\lambda$-Calculus}{The Safe Lambda-Calculus}}

\author[W. Blum, C.-H. L. Ong]{\texorpdfstring{\\ William Blum and C.-H. Luke Ong}{William Blum}}


\institute[University Of Oxford]{Oxford University Computing Laboratory}

\date{\small \color{red}{TLCA 2007}}


\begin{document}

\section{Title page}
  \frame{\titlepage}


%\section<presentation>*{Outline}
%\begin{frame}
%  \frametitle{Outline}
%  \tableofcontents[part=1]
%\end{frame}
%\AtBeginSection[] {
%   \begin{frame}<beamer>
%     \frametitle{Outline}
%     \tableofcontents[currentpart,currentsection]
%   \end{frame}
% }
%
%\part<presentation>{Main Talk}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Overview}
\frame{\frametitle{Overview}

\begin{itemize}
\item \highlight{Safety} is originally a syntactic restriction for higher-order grammars
with nice automata-theoretic characterization.
\item In the context of the $\lambda$-calculus it gives rise to the \highlight{Safe $\lambda$-calculus}.
\item The loss of expressivity can be characterized in terms of representable numeric functions.
\item The calculus has a ``succinct'' game-semantic model.
\end{itemize}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Outline for this talk}
\frame{
\frametitle{Outline for this talk}
\begin{enumerate}
\item The safety restriction for higher-order grammars
\item The safe $\lambda$-calculus
\item Expressivity
\item Game-semantic characterisation
\item Safe PCF, Safe IA
\end{enumerate}
}

\section{Higher-order grammars}
\frame{\frametitle{Higher-order grammars}
\emph{Notation for types:} $A_1 \rightarrow (A_2 \rightarrow (\ldots (A_n \rightarrow o)) \ldots )$
is written $(A_1,A_2,\ldots, A_n,o)$.

\begin{itemize}

\item Higher-order grammars (Maslov, 1974) are used as generators of word languages, trees or graphs.

\item A \highlight{higher-grammar} is formally given by a tuple
$\langle \Sigma, \mathcal{N}, \mathcal{R}, \mathcal{S} \rangle$
(terminals, non-terminals, rewritting rules, starting symbol)

\item Example of a tree-generating order-2 grammar:
\begin{columns}
      \column{.3\textwidth}
$\begin{array}{rll}
  S & \rightarrow & H \, a\\
  H \, z^o & \rightarrow & F \, (g \,
  z)\\
  F \, \phi^{(o, o)} & \rightarrow & \phi \, (\phi \, (F \, h))\\
\end{array}$
      \column{.3\textwidth}
\begin{tikzpicture}[baseline=(root.base),level distance=5mm,inner ysep=0.5mm,sibling distance=10mm]
 \node (root) {$g$}
    child {node {$a$}}
    child {node {$g$}
        child { node{$a$} }
        child { node{$h$}
                child { node{$h$}
                        child { node{$\vdots$} }
                }
        }
    } ;
\end{tikzpicture}
\end{columns}
Non-terminals: $S :o$, $H:(o,o)$ and $F:((o,o),o)$. Terminals: $a:o$ and $g,h:(o,o)$.
\end{itemize}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Safety Restriction}
\frame{\frametitle{The Safety Restriction}
\begin{itemize}
\item First appeared under the name ``restriction of derived types'' in ``IO and OI Hierarchies'' by W. Damm, TCS 1982
\item It is a \highlight{syntactic restriction} for higher-order grammars that constrains the occurrences
of the variables in the grammar equations according to their orders.
\pause

\item $(A_1, \cdots, A_n, o)$ is \highlight{homogeneous} if
$A_1$, \ldots, $A_n$ are and $\ord{A_1} \geq \ord{A_2}\geq \cdots \geq \ord{A_n}$.
\end{itemize}

\begin{definition}[Knapik, Niwi\'nski and Urzyczyn (2001-2002)]
\noindent All types are assumed to be \emph{homogeneous}.

  An order $k > 0$ term is \emph{unsafe} if it contains an
  occurrence of a parameter of order strictly less than $k$.
  An unsafe subterm $t$ of $t'$ occurs in \emph{safe position}
  if it is in operator position ($t' = \cdots (ts) \cdots$).
  A grammar is \highlight{safe} if at the right-hand side of any production
  all unsafe subterms occur in safe positions.
\end{definition}
}

\section{Some Results On Safety}
\frame{\frametitle{Some Results On Safety}
\begin{description}
\item [Damm82] For generating word languages, order-$n$ safe grammars are
equivalent to order-$n$ pushdown automata.

\item [KNU02] Generalization of Damm's result to
\emph{tree generating} safe grammars/PDAs.

\item [KNU02] The Monadic Second Order (MSO) model checking problem for trees generated by
    \highlight{safe} higher-order grammars of any order is decidable.

\item [Ong06] But anyway, KNU02 result's is also true for unsafe grammars...

\item [Caucal02] Graphs generated by safe grammars have a decidable MSO theory.

\item [HMOS06] Caucal's result does not extend to unsafe grammars.
However deciding $\mu$-calculus theories is $n$-EXPTIME complete.

\item [AdMO04] Proposed a notion of safety for the $\lambda$-calculus (unpublished).

\end{description}


\note{
\begin{itemize}
\item nPDA = finite state machines + order n stack
\item For words: 1PDA recognizes context-free language.
        and 0PDA = recognizes regular language.
\item MSO is very expressive: more than the modal mu-calculus (into which LTL CTL CTL* can be embedded.
But over trees, MSO and modal mu-calculus are equi-expressive.)
\end{itemize}
}
}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Simply Typed \texorpdfstring{$\lambda$}{Lambda}-Calculus}
\frame{\frametitle{Simply Typed $\lambda$-Calculus}
\begin{itemize}
\item \highlight{Simple types} $A := o\ |\ A \rightarrow A$.
%We write $(A_1,\ldots, A_n)$ for $A_1\rightarrow \ldots \rightarrow A_n$.

\item The \highlight{order} of a type is given by $\textsf{order}(o) = 0$,
$\textsf{order}(A \rightarrow B) = \max(\textsf{order}(A) + 1, \textsf{order}(B))$.

\item Jugdements of the form $ \Gamma \vdash M : T $ where $\Gamma$ is the context, $M$ is the term and $T$ is the type:
$$ \rulename{var} \   \rulef{}{x : A\vdash x : A}
\qquad  \rulename{wk} \   \rulef{\Gamma \vdash M : A}{\Delta \vdash M : A} \ \Gamma \subset \Delta$$
$$ \rulename{app} \  \rulef{\Gamma \vdash M : A \rightarrow B \quad \Gamma \vdash N : A }
                           {\Gamma  \vdash M N : B}
\quad \rulename{abs} \   \rulef{\Gamma, x : A \vdash M : B}
                                {\Gamma  \vdash \lambda x^A. M : A \rightarrow B}$$

\item Example: $f:o\rightarrow o\rightarrow o, x:o \vdash (\lambda \varphi^{o \rightarrow o} x^o . \varphi\ x) (f\ x)$

\item A single rule: \highlight{$\beta$-reduction}. e.g. $(\lambda x. M) N \betared M [N/x]$
\end{itemize}
}




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Safe \texorpdfstring{$\lambda$}{Lambda}-Calculus}
\begin{frame} \frametitle{The Safe $\lambda$-Calculus}

\begin{block}{The formation rules}
$$ \rulename{var} \   \rulef{}{x : A\vdash_s x : A}
\qquad  \rulename{wk} \   \rulef{\Gamma \vdash_s M : A}{\Delta \vdash_s
M : A} \ \Gamma \subset \Delta$$
$$ \rulename{app} \  \rulef{\Gamma \vdash M : (A_1,\ldots,A_l,B)
                                        \quad \Gamma \vdash_s N_1 : A_1
                                        \quad \ldots \quad \Gamma \vdash_s N_l : A_l  }
                                   {\Gamma  \vdash_s M N_1 \ldots N_l : B}$$
\hfill with the side-condition $\textcolor{red}{\forall y \in \Gamma
: \ord{y} \geq \ord{B}}$
$$ \rulename{abs} \   \rulef{\Gamma, x_1:A_1 \ldots x_n : A_n \vdash_s M : B}
                                   {\Gamma  \vdash_s \lambda x_1:A_1 \ldots x_n : A_n . M : A_1 \rightarrow \ldots \rightarrow A_n \rightarrow B}$$
\hfill with the side-condition $\textcolor{red}{\forall y \in \Gamma
: \ord{y} \geq \ord{A_1 \rightarrow \ldots \rightarrow A_n
\rightarrow B}}$
\end{block}

\begin{lemma}
  If $\Gamma \vdash_s M : A$ then every free variable in $M$ has order at least $\ord{A}$.
\end{lemma}

\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Variable Capture}
\frame{\frametitle{Variable Capture} \highlight{The usual
``problem'' in $\lambda$-calculus}: avoid \alert{variable capture}
when performing substitution: $ (\lambda x . (\lambda y . x)) y
\betared (\lambda \underline{y} . x) [\underline{y}/x] \neq \lambda
y . y$\pause
\begin{enumerate}
\item \highlight{Standard solution}: Barendregt's convention. Variables are renamed so that free variables and bound variables have different names.
Eg. $(\lambda x . (\lambda y . x)) y$ becomes $(\lambda x . (\lambda
z . x)) y$ which reduces to $(\lambda z . x) [y/x] = \lambda z . y$
\pause

\alert{Drawback:} requires to have access to an unbounded supply of names to perform
a given sequence of $\beta$-reductions.
\note{Drawback 1, eg. $(\lambda x_1 \ldots x_n . (\lambda y_1 \ldots y_n . x_1 \ldots x_n)) y_1 \ldots y_n$}
\pause

\item \highlight{Another solution}: use the $\lambda$-calculus \`a la de Brujin where variable binding is specified by an index instead of a name.
Variable renaming then becomes unnecessary.
\pause

\alert{Drawback:} the conversion to nameless de Brujin $\lambda$-terms requires an unbounded supply of indices.
\end{enumerate}

\pause
\begin{block}{Property}
In the Safe $\lambda$-calculus there is no need to rename variables when
performing substitution.
\end{block}

}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Examples}
\frame{\frametitle{Examples}
\begin{enumerate}
\item
Contracting the $\beta$-redex in the following term
$$f:o\rightarrow o\rightarrow o, x:o \vdash (\lambda \varphi^{o \rightarrow o} x^o . \varphi\ x) (\temporal<2>{f\ x}{\underline{f\ \alert{x}}}{\underline{f\ \alert{x}}})$$
leads to variable capture:
$$(\lambda \varphi x . \varphi\ x) (f\ x) \not\betared (\lambda \alert{x} . (f\ \alert{x}) x).$$
\pause
Hence the term is \alert{unsafe}. Indeed, $\ord{x} = 0 \leq 1 = \ord{f\ x}$.
\pause

\item The term $(\lambda \varphi^{o \rightarrow o} x^o . \varphi\ x) (\lambda y^o . y )$ is safe.
\pause

\item Safety does not capture ``variable-renaming uselessness''. E.g.~the unsafe term
$\lambda y^o z^o. (\lambda x^o .y) z$ can be contracted using capture-permitting substitution.
\pause

\item Up to order $2$, $\beta$-normal terms are always safe.
\pause

\item \highlight{Kierstead terms} $\lambda f^{((o,o),o)} . f (\lambda x^o . f (\lambda y^o . y
))$ is safe but $\lambda f^{((o,o),o)} . f (\lambda x^o . f (\underline{\lambda y^o .x}))$ is unsafe.

\end{enumerate}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Transformations preserving safety}
\frame{\frametitle{Transformations preserving safety}
\begin{itemize}
  \item Substitution preserves safety.
  \pause
  \item $\beta$-reduction does not preserve safety:
Take  $w,x,y,z : o$ and $f : (o,o,o)$.
The safe term $(\lambda x y . f\ x\ y) z\ w$
$\beta$-reduces to the unsafe term $(\underline{\lambda y . f\ z\ y}) w$
which in turns reduces to the safe term $f\ z\ w$.
  \pause

  \item Safe $\beta$-reduction: reduces simultaneously as many $\beta$-redexes
as needed in order to reach a safe term.
  \item Safe $\beta$-reduction preserves safety.
  \pause
  \item $\eta$-reduction preserves safety.
  \pause
  \item $\eta$-expansion \alert{does not} preserve safety. E.g.~
  $\vdash_s \lambda y^o z^o . y : (o,o,o)$ but
$\not \vdash_s \lambda x^o . (\lambda y^o z^o . y) x : (o,o,o)$.
  \pause
  \item $\eta$-long normal expansion preserves safety.
\end{itemize}

}


\section{Expressivity}
\frame{ \frametitle{Expressivity}

Safety is a strong constraint but it
is still unclear how it restricts expressivity:
\begin{itemize}
\item de Miranda showed that at order 2 for word languages,
non-determinism palliates the loss of expressivity.
It is unknown if this extends to higher orders.

\item For tree-generating grammars: Urzyczyn conjectured that
safety is a proper constraint i.e.~that there is a tree which is intrinsically unsafe.
He proposed a possible counter-example.

\item For graphs, HMOS06's undecidability result implies that safety restricts expressivity.

\item For simply-typed terms: ...

\end{itemize}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Numerical functions}
\frame{ \frametitle{Numerical functions}

Church Encoding: for $n\in\nat$, $\overline{n} = \lambda s z. s^n z$
of type $I = (o\rightarrow o)\rightarrow o\rightarrow o$.

\begin{theorem}[Schwichtenberg 1976]
The numeric functions representable by simply-typed terms of type $I\rightarrow \ldots \rightarrow I$ are
exactly the multivariate polynomials extended with the conditional function:
$$ cond(t,x,y) =
\left\{ \begin{array}{ll} x, & \mbox{if $t=0$} \\
y, & \mbox{if $t = n+1$}\ .
\end{array} \right.
$$
\end{theorem}
$cond$ is represented by the term $C = \lambda F G H \alpha x . H (
\temporal<2>{\lambda y . G \alpha x}
{\underline{\lambda y . G \alpha x}}{\underline{\lambda y . G \alpha x}}
 ) (F \alpha x)$.
\pause
\begin{theorem}
Functions representable  by safe $\lambda$-expressions of type $I\rightarrow \ldots \rightarrow I$  are
exactly the multivariate polynomials.
\end{theorem}
So $cond$ is not representable in the Safe $\lambda$-calculus and $C$ is unsafe.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%\section{Game Semantics}
%%%%\frame{ \frametitle{Game Semantics}
%%%%Let $\vdash M:T$ be a pure simply typed term.
%%%%
%%%%\begin{itemize}
%%%%\item \highlight{Game-semantics} provides a model of $\lambda$-calculus.
%%%%$M$ is denoted by a strategy $\sem{M}$ on a 2-player game induced by $T$.
%%%%\item A \highlight{strategy} is represented by a set of sequences of moves together with \highlight{links}: each move points
%%%%to a preceding move.
%%%%\pause
%%%%\item \textcolor{DarkGreen}{Computation tree} = canonical tree representation of a term.
%%%%\item \textcolor{DarkGreen}{Traversals $\travset(M)$ } = sequences of nodes with links respecting some formation rules.
%%%%\end{itemize}
%%%%\pause
%%%%
%%%%\begin{block}{The Correspondence Theorem}
%%%%The game semantics of a term can be represented on the computation
%%%%tree:
%%%%$$ \textcolor{DarkGreen}{\travset(M)} \cong \textcolor{blue}{\intersem{M}} $$
%%%%$$ \textcolor{DarkGreen}{Reduction(\travset(M))} \cong \textcolor{blue}{\sem{M}}$$
%%%%where $\textcolor{blue}{\intersem{M}}$ is the revealed game-semantic denotion (i.e. internal moves are uncovered).
%%%%\end{block}
%%%%}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Game semantics}
\frame{\frametitle{Game semantics}
 Model of programming languages based on games (Abramsky et al.; Hyland and Ong; Nickau)
\begin{itemize}
\item 2 players: \highlight{O}pponnent (system) and \highlight{P}roponent (program)
\item The term type induces an \highlight{arena} defining the possible moves
$\sem{\nat} = \begin{tikzpicture}[baseline=(root.base),level distance=7mm,inner ysep=0.5mm,sibling distance=5mm]
 \node (root) {$q$}
    child {node {$\scriptstyle 0$}}
    child {node {$\scriptstyle 1$}}
    child {node {$\scriptstyle \ldots$}}
;
\end{tikzpicture}$
\hspace{2cm}
$\sem{\nat \rightarrow \nat} = \begin{tikzpicture}[baseline=(root.base),level distance=7mm,inner ysep=0.5mm,sibling distance=5mm]
 \node (root) {$q^0$}
    child{
      node{$q^1$}
      child{node {$\scriptstyle 0$} }
      child{node {$\scriptstyle 1$} }
      child {node {$\scriptstyle \ldots$}}
    }
    child {node {$\scriptstyle 0$}}
    child {node {$\scriptstyle 1$}}
    child {node {$\scriptstyle \ldots$}}
;
\end{tikzpicture}$
\item \highlight{Play} = justified sequence of moves played alternatively by O and P with justification pointers.
\item \highlight{Strategy for P} = prefix-closed set of plays. $s  a  b$ in the strategy means that
P should respond $b$ when O plays $a$ in position $s$.
\item The \highlight{denotation} of a term $M$, written $\sem{M}$, is a strategy for P.
\item $\sem{ 7 : \nat} = \{ \epsilon, q, q\ 7 \}$\\
$\sem{ \pcfsucc : \nat \rightarrow \nat} = Pref( \{ q^0 q^1 n ( n+1)
\ | \ n \in \nat \} )$
\item Compositionality: $\sem{ \pcfsucc\  7} = \sem{ \pcfsucc } ; \sem{7}$
\end{itemize}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Game-semantic Characterisation of Safety}
\frame{ \frametitle{Game-semantic Characterisation of Safety}

The variable binding restriction imposed by the safety constraint
implies:
    \begin{block}{Theorem}
    \begin{itemize}
    \item Safe terms are denoted by \highlight{P-incrementally justified strategies}: each
    P-move $m$ points to the last O-move in \highlight{the
    P-view} with order $> \ord{m}$.
    \item Reciprocally, if a \emph{closed} term is denoted by a \highlight{P-incrementally justified strategy}
     then its $\eta$-long $\beta$-normal form is safe.
    \note{``closed'' is necessary: take $\lambda x y . x$ and $\lambda y . x$.}
    \end{itemize}
    \end{block}
\pause

\begin{block}{Corollary}
Justification pointers attached to P-moves are redundant in the game-semantics of safe
terms.
\end{block}
}


\section{Safe PCF}
\frame{ \frametitle{Safe PCF}

\begin{itemize}
\item \highlight{PCF} = $\lambda^{\rightarrow}$ with base type $\nat$ +  successor, predecessor, conditional + Y combinator

\item \highlight{Safe PCF} = Safe fragment of PCF
\end{itemize}

\begin{block}{Proposition}
  Safe PCF terms are denoted by P-i.j. strategies.
\end{block}

The first fully-abstract models of PCF were based on game semantics
(Abramsky et al., Hyland and Ong, Nickau).


\highlight{Question:} Is the game model of Safe PCF fully-abstract?
 }



\section{Idealized Algol : Open problem}
\frame{ \frametitle{Idealized Algol (IA) : Open problem }
\begin{itemize}
\item \highlight{IA} = PCF + imperative features (sequencing, variable dereferencing/assignment, block structure)
\item Introduced by John Reynolds, 1997.

\item $IA_i+Y_j$: fragment of IA with finite base type, terms of order
$\leq i$, recursion limited to order $j$

\item Safe IA = the Safe fragment of IA
\end{itemize}

\highlight{Question}: Is observational equivalence decidable for
Safe IA?

Abramsky and McCusker showed that two IA terms are equivalent if and
only if the set of complete plays of their game denotations is the
same [Abramsky,McCusker].

\begin{itemize}
\item $IA_2$: the set of complete plays is regular [Ghica].

\item $IA_3 + Y_0$: DPDA definable [Ong].

\item $IA_3 + while$: Visibly Pushdown Automaton definable [Murawski].
\end{itemize}
Hence observational equivalence is decidable for all these
fragments.

At order 4, however observational equivalence is undecidable
[Murawski]. This is partially due to the fact that pointers cannot
be encoded succinctly.

}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion and Future Works}
\frame{ \frametitle{Conclusion and Future Works}

\highlight{Conclusion:}

Safety is a syntactic constraint with interesting algorithmic and
game-semantic properties.

\highlight{Future works:}
\begin{itemize}
\item Can we obtain a fully abstract model of Safe PCF (with respect to safe contexts)?
\item Complexity classes characterised with the Safe $\lambda$-calculus?
\item Safe Idealized Algol: is contextual equivalence decidable
for some finitary fragment (e.g.~Safe IA$_4$) (with respect to all/safe contexts) ?
\end{itemize}


\highlight{Related works:}
\begin{itemize}
\item Jolie G. de Miranda's thesis on safe/unsafe grammars.
\item Ong introduced computation trees in LICS2006 to prove decidability of MSO theory on infinite trees
generated by higher-order grammars (whether safe or not).
%\item Stirling recently proved decidability of higher-order pattern matching with a game-semantic approach
%relying on equivalent notions of computation tree and traversal.
\end{itemize}
}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Bibliography}

\end{document}
\endinput

\begin{frame} \frametitle<presentation>{Bibliography}

%  \begin{thebibliography}{10}
  \beamertemplatearticlebibitems
    \bibitem{abramsky:game-semantics-tutorial}
    Samson Abramsky and Guy McCusker.
    \newblock Game semantics, Lecture notes.
    \newblock In {\em Proceedings of the 1997 Marktoberdorf Summer School}. Springer-Verlag, 1998.

    \bibitem{safety-mirlong2004}
    Klaus Aehlig, Jolie~G. de~Miranda, and C.-H.~Luke Ong.
    \newblock Safety is not a restriction at level 2 for string languages.
    \newblock Technical report. University of Oxford, 2004.

    \bibitem{OngLics2006}
    C.-H.~Luke Ong.
    \newblock On model-checking trees generated by higher-order recursion schemes.
    \newblock In {\em Proceedings of LICS.} Computer Society Press, 2006.

%    \bibitem{DBLP:conf/icalp/Stirling06}
%    Colin Stirling
%    \newblock A Game-Theoretic Approach to Deciding Higher-Order Matching.
%    \newblock In {\em Proceedings of ICALP.} Springer, 2006.

%  \end{thebibliography}
\end{frame}

\end{document}
