\input{lmcs.pre}

%% due to the dependence on amsart.cls, \begin{document} has to occur
%% BEFORE the title and author information:
\begin{document}

\title[The Safe Lambda Calculus]{The Safe Lambda Calculus \\ {\small \it version of \today}}

\author[W.Blum and C.-H. L.Ong]{William Blum}   %required
\address{Oxford University Computing Laboratory, Oxford, UK}    %required
% Wolfson Building, Parks Road, Oxford OX1 3QD, ENGLAND
\email{william.blum@comlab.ox.ac.uk}  %optional
%\thanks{thanks 1, optional.}   %optional

\author[]{C.-H.~Luke~Ong}   %optional
\address{Oxford University Computing Laboratory, Oxford, UK} %optional ; addresses should be duplicated when authors share an  affiliation
% Wolfson Building, Parks Road, Oxford OX1 3QD, ENGLAND
\email{luke.ong@comlab.ox.ac.uk}  %optional ; ditto for email addresses
\thanks{thanks 2, optional.}    %optional


%% etc.

%% required for running head on odd and even pages, use suitable
%% abbreviations in case of long titles and many authors:

%% mandatory lists of keywords and classifications:
\keywords{MANDATORY list of keywords}
\subjclass{MANDATORY list of acm classifications}

% OPTIONAL comment concerning the title, \eg, if a variant
%or an extended abstract of the paper has appeared elsewhere
\titlecomment{Some of the results presented here were first published in TLCA proceedings \cite{blumong:safelambdacalculus}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% the abstract has to PRECEED the command \maketitle:
%% be sure not to issue the \maketitle command twice!

\begin{abstract}
  Safety is a syntactic condition of higher-order grammars that
  constrains occurrences of variables in the production rules
  according to their type-theoretic order. In this paper, we introduce
  the \emph{safe lambda calculus}, which is obtained by transposing
  (and generalizing) the safety condition to the setting of the
  simply-typed lambda calculus. In contrast to the original definition
  of safety, our calculus does not constrain types (to be
  homogeneous). We show that in the safe lambda calculus, there is no
  need to rename bound variables when performing substitution, as
  variable capture is guaranteed not to happen.  We also propose an
  adequate notion of $\beta$-reduction that preserves safety.  In the
  same vein as Schwichtenberg's 1976 characterization of the
  simply-typed lambda calculus, we show that the numeric functions
  representable in the safe lambda calculus are exactly the
  multivariate polynomials; thus conditional is not definable. We
  also give a characterization of representable word functions.
  We then study the complexity of deciding beta-eta equality of two safe simply-typed term and show that this
  problem PSPACE-hard.
  Finally we give a game-semantic analysis of safety: We show that
  safe terms are denoted by \emph{P-incrementally justified
    strategies}. Consequently pointers in the game semantics of safe
  $\lambda$-terms are only necessary from order 4 onwards.
\end{abstract}

    \maketitle              % typeset the title of the contribution

    \include{safelambda}

    \include{safe_expressivity}
    \include{safe_complexity}

    \include{safe_gamesem}

    \include{further}

    \bibliographystyle{abbrv} % {nat} % {splncs}
    \bibliography{../bib/dphil-all}

    \section*{Appendix A -- Computation tree, traversals and correspondence}
    \label{sec:correspondence}
    \include{corresp_overview}

\end{document}
