 \ifx\incrcompilation\undefined \input preamble.tex \fi
% 
%\includeonly{chap_previouswork,chap_gamesem,safe_background,safe_homog,safe_nonhomog,safe_gamesem2,chap_further}
%\includeonly{safe_gamesem2} 
                          
 
\author{William Blum}
\title{Transfer thesis}
\institution{Oxford University Computing Laboratory}

\begin{document}
\maketitle \tableofcontents
          
\chapapp{}
\part{Academic activities} %Previous work and plan of proposed work}
\include{chap_previouswork}

\setcounter{chapter}{0}
\chapapp{Chapter}
\part{Summary of work so far}

The first chapter of this part is devoted to the presentation of the
basics and main results of game semantics. The categorical
interpretation of game semantics is presented as well as the full
abstraction result for \pcf. We also give a brief summary of the
main results in algorithmic game semantics. There is no personal
contribution in this chapter.

In the second chapter we present the \emph{Safe $\lambda$-Calculus}.
Originally, \emph{safety} has been introduced as a syntactical
restriction on higher-order grammars in order to show a decidability
result about MSO theory of infinite trees \citep{KNU02}. In
\cite{safety-mirlong2004}, Aehlig, de Miranda and Ong  proposed an
adaptation of the safety restriction to the $\lambda$-calculus. This
restriction gives rise to the Safe $\lambda$-Calculus. We first
present this calculus and then give a more general definition which
does not make any assumption on the types of the terms.

In the third chapter, following ideas described in
\cite{OngLics2006}, we introduce the notions of computation tree of
a simply-typed term and traversal over a computation tree. We prove
a theorem showing a correspondence between traversals of the
computation tree and the game semantics of a term. Based on that
correspondence, we give a characterisation of the game semantics of
safe terms by a property called ``incremental justification''. In
incrementally-justified strategies, pointers are superfluous (i.e.
they can be recovered uniquely from the underlying sequence of
moves). This simplification of the game semantics suggests some
potential applications in algorithmic game semantics. We finish the
chapter by extending the result to Safe \pcf\ and by giving the key
elements for an extension to full Safe Idealized Algol.


% first chapter
\include{chap_gamesem}

% second chapter
\include{chap_safe_background}
\include{chap_safe_homog}
\include{chap_safe_nonhomog}
% \include{safe_nonhomog_va}

% third chapter
\include{chap_comptrav}

% fourth chapter
\include{chap_gsemsafety}

% fifth chapter
\include{chap_further}

\bibliographystyle{plain}
\bibliography{../../bib/gamesem,../../bib/modelchecking,../../bib/proganalys,../../bib/higherorder}

%adds the bibliography to the table of contents
\addcontentsline{toc}{chapter}
         {\protect\numberline{Bibliography\hspace{-96pt}}}
\end{document}
