
\subsection{Pubs}

The following is a brief description of some of the main pubs in Oxford, along with their map reference on the \textit{Oxford by Night} map provided.  Pubs usually finish serving food about 1900 to 2030.  Standard pub food includes burgers, lasagne, steak, sausage and mash, chicken tikka masala and fish and chips.

\begin{description}


\item[Duke of Gloucester] Map reference D2.  FILL ME IN!

\item[Lamb and Flag] Map reference D3.  Food \pounds 3.50-5

\item[The Goose] By the Gloucester Arms.  Map reference D4.  Has a large beer garden and the occasional barbeque.  Food roughly \pounds 5, with meal deals (with drink) starting from \pounds 3.

\item[Old Orleans] On George Street.  Map reference D5.  Mississippi themed pub/restaurant, serves cajun and creole cooking.  Burgers \pounds 7-10, main courses \pounds 9-13.

\item[Three Goats Heads] By the Oxford Union.  Map reference D5.  Has an unusual selection of drinks.  Food \pounds 5-6.

\item[O'Neils] On George Street.  Map reference D5.  Chain Irish theme pub.  Burgers \pounds 5, main courses \pounds 6-7.


\item[The Chequers] Map reference E6.  Has three pool tables and two quiz machines.  Food \pounds 6.

\item[The Mitre] Map reference E5.  Food \pounds 8-12.

\item[The Pub Oxford] By the Cricketer's Arms.  Map reference I7.  Lots of games machines including a games console.  Food \pounds 5.

\item[Head of the River] Map reference E8.  Large beer garden on the river.  Punting nearby.  Food \pounds 8-9.


\end{description}




\subsection{Bars}

Below is a guide to some of the bars in Oxford.  Not all of them serve food.  Map references refer to the \textit{Oxford by Night} map.

\begin{description}

\item[Freuds] Map reference C2.  A converted church with live music and local art.  Serves calzone, bagels and ciabattas for \pounds 5-6.

\item[Baby] Map reference C2.  Mediterranean bar, serves cocktails.  Food \pounds 13.

\item[Copa] On George Street.  Map reference D5.  Serves a range of continental beers.  Food \pounds 7-8.

\item[All Bar One] 124 High Street.  Wine bar aimed at over 21s.  Sharing platters for three \pounds 11.  Burgers, salads, steaks \pounds 7-10.

\item[Quod] Map reference F5. Bar and grill.  Large sun terrace.  Food \pounds 8-10.

\item[Thirst] Map reference B5.  Cheap cocktails, open late.  

\end{description}

\subsection{Lunch -- Market Square}

Market Square lies behind the Odeon Cinema on George Street and Gloucester Green Coach Station.  Map reference D5.

\begin{description}

\item[Noodle Bar] Cheap noodles.  \pounds 4-5.

\item[Cleo's Cafe] Sweet and savoury crepes.  Free chocolate crepe with every savoury crepe.  \pounds 4.50.

\item[Harveys] Made to order hot sandwiches.  \pounds 2-3.

\item[Cafe Piazza] Sanwiches, bagels, baguettes and ciabattas for \pounds 3-4.
  

\end{description}


\subsection{Lunch -- Cornmarket Street}

A brief description of the lunch places near Cornmarket Street.  Map references are given where appropriate, and refer to the \textit{Oxford by Day} map.  Cornmarket Street can be found at E5.

\begin{description}

\item[Sainsbury's Local] On Magdelen Street. Map reference E4.  Sells pre-packaged sandwiches for \pounds 1.50-3.

\item[Fast Food] The standard fast food places are all represented on Cornmarket Street (KFC, Burger King, McDonalds).  Burger King and KFC are on the George Street side, whilst McDonalds is towards the High Street.  Meals (burger, fries, drink) are roughly \pounds 4.

\item[St Michael's Street]  St Michaels Street is just off Cornmarket Street.  There are two lunch places.  The Nosebag serves salads, baked potatoes, soup, etc. for \pounds 4-5.  Meltz serves sandwiches, salads and pasta for \pounds 5-7.

\item[Ship Street] Ship street is opposite St Michael's Street.  The News Cafe sells salads, pasta, bagels for \pounds 5-8.50.  Heroes sell sandwiches made to order for \pounds 3-4 (takeaway, plus 20\% eat-in).

\item[West Cornwall Pasty] Near the High Street.  Sells pre-made cornish pasties for \pounds 2.50.  

\item[Golden Gate Covered Market] Entrance near West Cornwall Pasty.  There are many cafes within the market, plus Pizza Express --- which sells pizzas for \pounds 6-8 and salads and pasta for \pounds 7-8 --- and Cafe Puccino --- which sells salads, paninis, ciabattas, baguettes, baked potatoes and tapas for roughly \pounds 6.  

\item[Pret] Near the High Street.  Sells packaged sandwiches (more interesting than Sainsbury's) for \pounds 2-3.

\end{description}

\subsection{Dinners -- Cornmarket Street / High Street / St Clements}

As well as the pubs and bars, there are also many restaurants in Oxford.

\begin{description}

\item[Pizza Express] Map reference E5.  Sells pizzas, pasta and salads for \pounds 6-8.  Not as cheap and cheerful as Pizza Hut.

\item[Pizza Hut] Opposite The Chequers.  Map reference E6.  Pizzas \pounds 5.50 for an individual.  On Tuesdays: eat as much as you like for \pounds 7 (includes drink).

\item[Fishers] Map reference J6.  Sells seafood.  Main courses \pounds 12-15.

\item[Pink Giraffe] Map reference J6.  Chinese restaurant.  \pounds 8-9 for a meal.

\item[Thai Orchid] Map reference J6.  Thai restaurant.  \pounds 9-10 for a meal.

\end{description}


\subsection{Nighttime Entertainment}

Below is a description of some of the cinemas/theatre/etc. in Oxford.  Please check the websites for what's on information.  Map references refer to \textit{Oxford by Night}.

\begin{description}

\item[Odeon Cinemas] Mainstream cinema.  Split between George Street (map reference D5) and Magdalen Street (map reference D4).  

URL: \verb+http://www.odeon.co.uk/+ (go to ``Times and Booking (text version)'' if Flash doesn't work.)

\item[Phoenix Picturehouse] Map reference B1.  More arthouse, but not completely obscure.  

URL: \verb+http://www.picturehouses.co.uk/site/cinemas/Oxford/local.htm+

\item[Ultimate Picture Palace] Map reference J7.  No frills cinema showing films from all over the world. Also fairly cheap.  

URL: \verb+http://www.ultimatepicturepalace.co.uk/+

\item[New Theatre] Map reference D5.  Lots of shows and musicals.  

URL: \verb+http://www.thisistheatre.com/regional/oxfordtheatre.html+


\item[Oxford Playhouse] Map reference D4.  As the name suggests, the focus is on plays rather than shows. 

URL: \verb+http://www.oxfordplayhouse.com/+

\item[Jongleurs]  Map reference B5.  Stand-up comedy, unfortunately only on Thursdays, Fridays and Saturdays.

URL: \verb+http://www.jongleurs.com/+


\end{description}



