\documentclass[nocenter,sfbold]{thesis}

\chapapp{Chapter}

\input preambule.tex

%\includeonly{chap_previouswork,chap_gamesem,chap_safelambda,safe_gamesem2}
%\includeonly{safe_gamesem2}


\author{William Blum}
\title{Transfer thesis}
\institution{Oxford University Computing Laboratory}

\begin{document}
\maketitle \tableofcontents

\part{Previous work and plan of proposed work}

\include{chap_previouswork}


\part{Research proposal}

The first chapter of this part is devoted to the presentation of the
basics and main results of game semantics. The categorical
interpretation of game semantics is presented as well as the full
abstraction result for PCF. We also give a brief summary of the main
results in algorithmic game semantics. There is no personal
contribution in this chapter.

In the second chapter we present the \emph{Safe $\lambda$-calculus}.
Originally, \emph{safety} has been introduced as a syntactical
restriction on higher-order grammars and has been used to show a
decidability result about MSO theory of infinite trees
\citep{KNU02}. In \cite{safety-mirlong2004}, Aehlig, de Miranda and
Ong  proposed an adaptation of the safety restriction to the
$\lambda$-calculus. This restriction gives rise to the Safe
$\lambda$-calculus. We first present this calculus and then give a
more general definition which does not make any assumption on the
types of the terms.

Following ideas described in \cite{OngLics2006}, we then introduce
the notion of computation tree of a simply-type term and traversals
over a computation tree.

We then prove a theorem showing a correspondence between traversals
of the computation tree and the game semantics of a term. Based on
that correspondence, we give a characterization of the game
semantics of safe terms by what we call ``regular strategies''.
Regular strategies are algorithmically interesting in the sense that
pointers are superfluous (i.e. can be recovered uniquely from the
sequence of moves).

Finally we give some hints for a possible extension to Safe
Idealized Algol.

\include{chap_gamesem}

\include{chap_safelambda}

\include{safe_gamesem2}


\bibliographystyle{plainnat}
\bibliography{gamesem,modelchecking,proganalys,higherorder}

%adds the bibliography to the table of contents
\addcontentsline{toc}{chapter}
         {\protect\numberline{Bibliography\hspace{-96pt}}}
\end{document}
